---
title: "Waterfowl Breeding Population Survey For Wisconsin, 1973--`r params$year`"
author:  |
  | Drew Fowler^1^ and Taylor Finger^2^
  | *^1^Office of Applied Science*
  | *^2^Bureau of Wildlife Management*
  | *Wisconsin Department of Natural Resources*
date: "`r format(Sys.time(), '%B %Y')`"
fontsize: 12pt
geometry: margin = 1in
params: 
  year: 
bibliography: survey_citations.bib
csl: the-journal-of-wildlife-management
header-includes:
  - \usepackage{booktabs}
  - \usepackage{colortbl}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{makecell}
output: 
  pdf_document:
    toc: true
    number_sections: true
    toc_depth: 3
---

\newpage

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# automate some parts of report text with previous year, number of study years
previous_year <- read_csv(str_c(here::here('output'), '/', analysis_year, '/', 'ws_structured.csv')) %>%
  distinct(year) %>%
  slice_tail(n = 2) %>%
  slice(1) %>%
  pull()
missing_years <- 1 # due to covid
num_survey_years <- length(seq(1973, analysis_year, 1)) - missing_years
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# wetland summary statistics
# reported in text below
extract_wetland_statistics(region_code = 1)
extract_wetland_statistics(region_code = 2)
extract_wetland_statistics(region_code = 3)
extract_wetland_statistics(region_code = 4)

sec_results <- read_csv(str_c(here::here('output'), '/', analysis_year, '/', 'wetland_summary_region_1.csv'))
nhi_results <- read_csv(str_c(here::here('output'), '/', analysis_year, '/', 'wetland_summary_region_2.csv'))
nlo_results <-read_csv(str_c(here::here('output'), '/', analysis_year, '/', 'wetland_summary_region_3.csv'))
swd_results <- read_csv(str_c(here::here('output'), '/', analysis_year, '/', 'wetland_summary_region_4.csv'))
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
# state-space model estimates for waterfowl species
# reported in text below
mall_results <- extract_state_space_statistics(spp = 'mallard')
bwte_results <- extract_state_space_statistics(spp = 'blue-winged teal')
wodu_results <- extract_state_space_statistics(spp = 'wood duck')
other_results <- extract_state_space_statistics(spp = 'other ducks')
all_results <- extract_state_space_statistics(spp = 'total ducks')
cago_results <- extract_state_space_statistics(spp = 'canada goose')

# get frequency of species compromising 'other' ducks
comp_other_ducks <- read_csv(str_c(here::here('output'), '/', analysis_year, '/', 'comp_oducks_state', "_", analysis_year, '.csv'))
```

# Abstract

The `r params$year` Waterfowl Breeding Population Survey for Wisconsin was conducted April 26--May 7 and followed methods of the North American waterfowl population survey. The information from this survey is used as part of the overall survey of breeding waterfowl in North America as well as in making state-level waterfowl management decisions. This survey has been conducted annually since 1973. While this survey was not conducted in 2020 due to the COVID-19 pandemic, data on Wisconsin waterfowl breeding populations are best interpreted as trends viewed over several years rather than as year-to-year changes.  

Compared to the long-term average (1981--2010), fall and winter (October--February) precipitation levels were down 11% throughout most of the state and spring (March--May) precipitation was 12% below normal, resulting in below-average wetland conditions across the state. 

Total non-linear basins were `r sec_results$direction_previous_non_linear` `r sec_results$percent_change_previous_non_linear` from `r previous_year` in the Southeast Central stratum but still `r sec_results$direction_long_term_non_linear` `r sec_results$percent_change_long_term_non_linear` from the long-term, `r num_survey_years`-year mean. In the Northern High stratum, total non-linear basins were `r nhi_results$direction_previous_non_linear` `r nhi_results$percent_change_previous_non_linear` from `r previous_year` and `r nhi_results$direction_long_term_non_linear` `r nhi_results$percent_change_long_term_non_linear` from the long-term mean. Non-linear basins were `r nlo_results$direction_previous_non_linear` `r nlo_results$percent_change_previous_non_linear` from `r previous_year` in the Northern Low stratum but still `r nlo_results$direction_long_term_non_linear` `r nlo_results$percent_change_long_term_non_linear` from the long-term mean. Non-linear basins were `r swd_results$direction_previous_non_linear` `r swd_results$percent_change_previous_non_linear` from `r previous_year` in the Southwest Driftless stratum and `r swd_results$direction_long_term_non_linear` `r swd_results$percent_change_long_term_non_linear` compared to this region's long-term (`r length(seq(1997, params$year, 1)) - missing_years`-yr) mean. Total linear basins were: `r sec_results$direction_previous_linear` `r sec_results$percent_change_previous_linear` from `r previous_year` in the Southeast Central stratum and `r sec_results$direction_long_term_linear` `r sec_results$percent_change_long_term_linear` from the long-term mean, `r nhi_results$direction_previous_linear` `r nhi_results$percent_change_previous_linear` from `r previous_year` in the Northern High stratum and `r nhi_results$direction_long_term_linear` `r nhi_results$percent_change_long_term_linear` from the long-term mean, and `r nlo_results$direction_previous_linear` `r nlo_results$percent_change_previous_linear` from `r previous_year` in the Northern Low stratum and `r nlo_results$direction_long_term_linear` `r nlo_results$percent_change_long_term_linear` from the long-term mean. Total linear basins were `r swd_results$direction_previous_linear` `r swd_results$percent_change_previous_linear` from `r previous_year` in the Southwest Driftless stratum and `r swd_results$direction_long_term_linear` `r swd_results$percent_change_long_term_linear` from the long-term mean. 

The total breeding duck modeled population estimate of `r all_results$mean` is `r all_results$deviation_previous_year` `r all_results$direction_previous_year` than the `r params$year` modeled estimate of `r all_results$previous_year_mean` and is `r all_results$deviation_long_term_mean` `r all_results$direction_long_term_mean` than the long-term, `r num_survey_years`-year mean. Overall, the total duck population estimate for `r params$year` is higher than what we have experienced the last few years (2015--2019) and above the total duck numbers experienced in the prior 10 years. The `r params$year` mallard modeled breeding population estimate of `r mall_results$mean` is `r mall_results$deviation_previous_year` `r mall_results$direction_previous_year` than the `r params$year - 1` modeled estimate of `r mall_results$previous_year_mean` and is `r mall_results$deviation_long_term_mean` `r mall_results$direction_long_term_mean` than the long-term mean (`r num_survey_years` years). The `r params$year` blue-winged teal modeled breeding population estimate of `r bwte_results$mean` is `r bwte_results$deviation_previous_year` `r bwte_results$direction_previous_year` compared to the `r params$year - 1` modeled estimate but remains `r bwte_results$deviation_long_term_mean` `r bwte_results$direction_long_term_mean` than the long-term average (`r num_survey_years`-year mean). At `r wodu_results$mean`, the `r params$year` population estimate for wood ducks is `r wodu_results$deviation_previous_year` `r wodu_results$direction_previous_year` than the `r params$year - 1` modeled estimate of `r wodu_results$previous_year_mean` and `r wodu_results$deviation_long_term_mean` `r wodu_results$direction_long_term_mean` than the long-term mean. We should always consider the continental perspective for migratory birds when interpreting local surveys. Fortunately, in recent years the continental duck populations have been near all-time high estimates. However, it is clear that our continued commitment to protecting wetland and grassland habitat and regulating harvest impacts are both important to the future of Wisconsin breeding ducks. The statewide modeled breeding Canada goose population estimate of `r cago_results$mean` is `r cago_results$deviation_previous_year` `r cago_results$direction_previous_year` compared to the `r params$year - 1` modeled estimate and `r cago_results$deviation_long_term_mean` `r cago_results$direction_long_term_mean` than the long-term (`r num_survey_years`-year) mean. Water conditions were average to below average prior to and during the time of the survey. This, coupled with an early spring prompted us to initiate the survey earlier than we have in previous years. Following the survey in May, we saw below-average rainfall, which may negatively impacted wetland conditions for the brood-rearing period. Average to above-average Canada goose breeding numbers should provide quality Canada goose hunting opportunities this fall.

\newpage

# Introduction

Decisions regarding hunting season structure and harvest limits in waterfowl management are based in part upon spring breeding pair surveys. The U.S. Fish and Wildlife Service's (USFWS) Waterfowl Breeding Population and Habitat Survey has been conducted for `r length(seq(1955, params$year, 1))` years across the traditional survey area of north-central United States, Canada and Alaska. The Wisconsin Waterfowl Breeding Population Survey, which is modeled after the continental survey, has been conducted for `r num_survey_years` years and provides a long-term measure of waterfowl breeding trends in Wisconsin. These data are used at the national and state level for monitoring waterfowl populations and making management decisions. Wisconsin's breeding waterfowl survey data are included in the Waterfowl Population Status Report published annually by the USFWS on continental waterfowl populations. In addition, during a typical year mallard data from Wisconsin, Minnesota, and Michigan are combined with data from the traditional survey areas as a basis for the Adaptive Harvest Management Report that is used to establish federal waterfowl season frameworks, however due to pandemic related restrictions this will not occur in `r params$year`. At the state level, waterfowl breeding survey data are used to inform annual hunting season regulations, identify long-term changes in species populations, and evaluate the impact of habitat changes and management. This report provides a summary and analysis of the `r params$year` survey data in support of these efforts. 

# Methods

The breeding waterfowl survey in Wisconsin is a stratified double sampling scheme patterned after the North American Waterfowl Breeding Population Survey developed by the USFWS [@platte1987standard] but modified for local conditions [@march1973breeding]. Fifty-five east-west aerial transects, each 30 mi long and 1/4 mi-wide, were randomly selected in 1973 within three strata of Wisconsin: Southeast Central region (SEC), Northern High Density region (NHI) and Northern Low Density region (NLO), based on duck densities and habitat (Figure \ref{fig:transect_map}). These strata total 43,359 mi^2^. In 1997, eleven aerial transects were added within the unglaciated Southwest Driftless region (SWD) (12,311 mi^2^) and included in the statewide population estimates. This area was not surveyed prior to 1997 because of its low wetland density. 

The `r params$year` aerial survey was conducted in 14 days, (April 26--May 7) using a Cessna 182 aircraft, flying 80--85 mph at a maximum altitude of 200 ft. An observer recorded waterfowl data from each side of the plane, with the observer on the north side of the plane recording the number and type of wetland basins within a transect. Visibility correction factors (*R*) for waterfowl not seen by the aerial crew were determined by counting waterfowl from the ground on segments (10--15 mi long) of 26 transects and calculated as the ratio of ground counts to air counts. In this analysis, *R* was calculated independently for all priority waterfowl species pooling data from all 26 transects. When the coefficient of variation (CV) of *R* exceeded 20%, data from prior years were iteratively added until the CV value was less than 20%. 

The aerial surveys are generally initiated in the southern part of Wisconsin with the crews progressing to the northern transects. The ground surveys were done on 11 days that spanned April 27--May 7. All ground counts were completed within two days of the air survey. Twenty-five people from two agencies were involved in the planning and implementation of the survey. 

Survey results separate mallards, blue-winged teal, wood ducks and Canada geese. All other ducks are pooled into a category of "other ducks." By 2004, wood duck populations had increased to the point where we were able to estimate them as a separate group rather than as part of "other ducks." This is significant because it allows us to provide independent breeding population estimates for three of the four duck species that constitute the majority of Wisconsin's fall harvest. 

This survey was not originally designed for surveying Wisconsin's resident Canada goose population. If a survey were specifically designed for breeding Canada geese in Wisconsin, it would be conducted earlier in the spring because of their earlier nesting chronology. Canada goose counts and population estimates were not included in this report prior to 1986 because of the small sample size. However, aerial counts of geese increased steadily from the mid-1980s through the early 2000s, making survey estimates useful indices to population trends of breeding Canada geese. Human-goose conflicts resulting from a growing goose population increase the importance of tracking the population status of breeding geese in Wisconsin.

Lesser scaup and bufflehead are not included in the state duck population estimates because they rarely breed in Wisconsin and when counted are assumed to be in migration to more northern breeding areas. 

Observers counted occupied and unoccupied wetland types during the aerial survey. The determination of wetland type from the air is difficult to standardize when observers change over years. Pooling data into linear (streams and ditches) and nonlinear (types I--VIII) wetland groups resolves some of the typing problems. The same aerial observers are used for a minimum five-year period to limit problems with observer bias. 

New to this year's report and analysis we used a state-space model [@kery2012bayesian] to determine population estimates. First, we calculated species-specific abundance estimates in each region using the traditional formula developed by XXX:

\begin{equation}
{\text{\textit{N}}} = {\text{\textit{B}}} \times {\text{\textit{A}}} \times {\text{\textit{R}}}
\end{equation}

where *B* is the bird density per mi^2^, *A* is the area of the region, and *R* is the visibility correction factor. Second, we used these traditional annual abundance estimates from 1973--`r params$year` as data in a state-space model to provide modeled abundance estimates that better account for observation error that inherently occurs in the survey. As a result of the state-space model, we get annual mean abundance estimates and associated 95% credible intervals. These credible intervals tell us the range of abundance estimate values within which we can be 95% certain to contain the true population value. Because annual spring breeding survey estimates are an unknown integration of both true annual population change and observation error, we elected to model all annual estimates for each priority species using a state-space modeling approach. State-space models are hierarchical models that decompose an observed time series of counts into a process variation (true population size change) and an observation error component (survey count biases). State-space models offer at least two important advantages. First, modeled survey estimates smooth out drastic annual changes in raw survey estimates that are biologically unrealistic (e.g., mallard abundance changing from roughly 250,000 in 1999, to 450,000 in 2000, and then to 180,000 in 2001). Secondly, a Bayesian state-space model allows for estimate prediction, even when counts are unavailable. In light of missing data for 2020, modeled estimates are available to fill in the gap. Therefore, in the following summaries of priority breeding waterfowl estimates, we reference abundance and percent changes in light of the modeled survey estimates rather than the raw survey estimates, but we provide both in the tables and figures for comparison. 

# Results \& Discussion

## Survey timing and weather

In `r params$year`, we saw warmer temperatures than normal across Wisconsin, with temperatures slightly warmer in the southern half of the state. In combination with below-average rainfall throughout much of the state, this required that we conduct this survey earlier than we have the past several years. We continued to see average and above average temperatures across most of the state in early May and mid-May, respectively. We initiated the breeding waterfowl survey on April 26, which appeared to be well timed with regards to mallards across much of the state. As in the past, the survey was initiated in the southern part of Wisconsin, progressing northward to account for the differences in phenology from south to north. The timing of the breeding waterfowl survey is always a challenge because variables such as weather, bird species phenology and tree leaf-out all impact the timing, visibility and accuracy of the survey. Weather was not much of an issue leading up to the survey, with dry and warm conditions statewide and most of the lakes in northern Wisconsin free of ice by early April. Conditions were mostly dry throughout the duration of the survey, so there was little influence on the number of temporary wetlands, which we often see when there are significant rain events as the survey is being conducted. There were some concerns with leaf-out on transects in the north-central portions of the state. When observers encounter transects that have leaf-out conditions it is considerably more difficult to observe birds in forested and shrub wetlands. 

The crews encountered few problems with waterfowl in groups of four or more birds in `r params$year`. Ten flocks of groups with more than 10 mallards were encountered, including one group of 50 ducks and another group of 100. Groups of five or more individuals made up 5% of the wood ducks, 16% of the Canada geese, 25% of the mallards, 69% of the blue-winged teal and 38% of the "other ducks" of the total indicated birds for each species. Flocks of >19 birds are not included in breeding population estimates as they are considered migrants or non-breeders.

## Precipitation

Fall and winter (October--February) precipitation was down throughout most of the state by 11% compared to the normal average (1981--2010). Spring (March--May) precipitation statewide was 12% below normal. When migrating ducks arrived in Wisconsin this spring, the typical high-density duck nesting areas in southern and eastern Wisconsin had below-normal water conditions, which likely contributed to our decline in mallard numbers.

## Wetlands

We had average to below-average precipitation throughout the fall and into the winter of `r params$year - 1`--`r params$year`. Snowpack was average to below-average throughout much of the state. With average runoff from the melting snow and the average conditions going into the fall of 2020, we saw below-average wet conditions throughout March and April, which caused dry conditions across the entire state during the time of the survey. Conditions remained dry throughout much of May with some little rainfall in the latter half of May, resulting in below-average wetland conditions during the brood rearing period. During the `r params$year` survey we observed wetland conditions differing with wetter conditions in the north and drier conditions in the south. 

Total non-linear basins were `r sec_results$direction_previous_non_linear` `r sec_results$percent_change_previous_non_linear` from `r previous_year` in the SEC (Table \ref{tab:wetland_tab_sec}; Figure \ref{fig:wetland_abundance_sec}) but still `r sec_results$direction_long_term_non_linear` `r sec_results$percent_change_long_term_non_linear` from the long-term (`r num_survey_years`-year) mean, `r nhi_results$direction_previous_non_linear` `r nhi_results$percent_change_previous_non_linear` from `r previous_year` in the NHI (Table \ref{tab:wetland_tab_nhi}; Figure \ref{fig:wetland_abundance_nhi}) and `r nhi_results$direction_long_term_non_linear` `r nhi_results$percent_change_long_term_non_linear` from the long-term mean, and similar compared to `r previous_year` in the NLO (Table \ref{tab:wetland_tab_nlo}; Figure \ref{fig:wetland_abundance_nlo}) but still `r nlo_results$direction_long_term_non_linear` `r nlo_results$percent_change_long_term_non_linear` from the long-term mean. Non-linear basins were `r swd_results$direction_previous_non_linear` `r swd_results$percent_change_previous_non_linear` from `r previous_year` in the SWD (Table \ref{tab:wetland_tab_swd}; Figure \ref{fig:wetland_abundance_swd}) and `r swd_results$direction_long_term_non_linear` `r swd_results$percent_change_long_term_non_linear` compared to this region's long-term (23-yr) mean. Total linear basins were: `r sec_results$direction_previous_linear` `r sec_results$percent_change_previous_linear` from `r previous_year` in the SEC (Figure \ref{fig:wetland_abundance_sec}) and `r sec_results$direction_long_term_linear` `r sec_results$percent_change_long_term_linear` from the long-term mean, `r nhi_results$direction_previous_linear` `r nhi_results$percent_change_previous_linear` from `r previous_year` in the NHI (Figure \ref{fig:wetland_abundance_nhi}) and `r nhi_results$direction_long_term_linear` `r nhi_results$percent_change_long_term_linear` from the long-term mean, and `r nlo_results$direction_previous_linear` `r nlo_results$percent_change_previous_linear` from `r previous_year` in the NLO (Figure \ref{fig:wetland_abundance_nlo}) and `r nlo_results$direction_long_term_linear` `r nlo_results$percent_change_long_term_linear` from the long-term mean. Total linear basins were `r swd_results$direction_previous_linear` `r swd_results$percent_change_previous_linear` from `r previous_year` in the SWD (Figure \ref{fig:wetland_abundance_swd}) and `r swd_results$direction_long_term_linear` `r swd_results$percent_change_long_term_linear` from the long-term mean. With the lack of precipitation in May, average to below average conditions for breeding and brood-rearing habitat in Wisconsin is expected.

## Waterfowl population estimates

### Mallards

The `r params$year` modeled **mallard population estimate is `r mall_results$mean` (95% CI: [`r mall_results$lower`; `r mall_results$upper`])**. This estimate is `r mall_results$deviation_previous_year` `r mall_results$direction_previous_year` compared to the previous year's modeled estimate and `r mall_results$deviation_long_term_mean` `r mall_results$direction_long_term_mean` than the long-term, `r num_survey_years`-year mean (Figure \ref{fig:ssm_mall}). 

Among duck species, mallards are highly adaptable to annual weather variation and often begin nesting when conditions may not yet be ideal. The early initiation of mallard nesting in `r params$year` should allow ample time for re-nesting efforts that can make a considerable contribution to annual production. Breeding pair survey numbers combined with average to below-average wetland conditions in the key breeding areas suggest that we will have mallard production similar to or lower than the last 10-year period. As in previous years, the SEC still represented the largest portion of the breeding mallard population (48%) and was similar to that of `r previous_year` (Table \ref{tab:current_breeding_estimate_table}). Because of the importance of the SEC region to mallard production and the increasing land use pressures, a focus on land acquisition, habitat restoration and management along with private land management incentives such as those available through the federal Farm Bill programs continue to be a priority for the SEC. Waterfowl breeding population estimates are best interpreted from long-term trends rather than year-to-year variation. 

### Blue-winged Teal

In `r params$year`, warm March and April temperatures likely resulted in slightly earlier migration and waterfowl breeding cycles. As a result, the timing of the breeding waterfowl survey, which targets breeding mallards, may have occurred too late to capture peak detection of blue-winged teal. The `r params$year` modeled population estimate for **Blue-winged teal is `r bwte_results$mean` (95% CI: [`r bwte_results$lower`; `r bwte_results$upper`])**. This estimate is `r bwte_results$deviation_previous_year` higher compared to the previous year's modeled estimate but remains `r bwte_results$deviation_long_term_mean` `r bwte_results$direction_long_term_mean` than the long-term, `r num_survey_years`-year mean (Figure \ref{fig:ssm_bwte}). The best approach when evaluating the blue-winged teal survey data is to look at long-term trends and continental context. In the late 1970s and 1980s, there was a clear decline in Wisconsin breeding blue-winged teal numbers concurrent with declines at the continental level. Since then, Wisconsin blue-winged teal breeding numbers have remained low compared to historic levels while continental numbers have shown dramatic increases during the late 1990s and in the last 10 years. We remain concerned about long-term decreases in secure grassland and nesting cover, particularly with fewer Conservation Reserve Program lands enrolled in Wisconsin. However, blue-winged teal are also known for shifting breeding locations around the continent in response to wetland habitat conditions. Continental blue-winged teal breeding populations have reached record highs over the past 7-8 surveys with estimates of 6-8 million breeding blue-winged teal reflecting good wetland conditions and good production in the prairies of Canada and the U.S. The abundant blue-winged teal populations in the prairies provide large fall flights of teal, which compensate for declines in Wisconsin. During the regular duck hunting season in Wisconsin, about two thirds of our blue-winged teal harvest is supported by teal that breed outside of Wisconsin. 

### Wood Ducks

The `r params$year` population estimate for **Wood ducks is `r wodu_results$mean` (95% CI: [`r wodu_results$lower`; `r wodu_results$upper`])**. This estimate is `r wodu_results$deviation_previous_year` `r wodu_results$direction_previous_year` compared to the previous year and is `r wodu_results$deviation_long_term_mean` `r wodu_results$direction_long_term_mean` than the long-term, `r num_survey_years`-year mean (Figure \ref{fig:ssm_wodu}). A combination of survey timing, improved habitat statewide and dry conditions this year, which may have concentrated wood ducks into more visible bodies of water (compared to flooded forest-land), could have contributed to an increase in wood duck detection and thus higher estimates compared to what we have seen in the past. As always, we remind people that it is important to view these population estimates as trends rather than focusing on year to year changes. The breeding wood duck population showed significant gains in the 1980s and early 1990s and appears to be leveling off around 100,000 after peaking about 10 years ago but has shown an increasing trend over the past five years. Based on improved water conditions and our best interpretation of the survey results, we expect wood duck production in `r params$year` to be excellent and up from recent years.

### Other Ducks

The `r params$year` modeled population estimate for **"other ducks"** is **`r other_results$mean` (95% CI: [`r other_results$lower`; `r other_results$upper`])**. This estimate is `r other_results$deviation_previous_year` `r other_results$direction_previous_year` compared to the previous year and is `r other_results$deviation_long_term_mean` `r other_results$direction_long_term_mean` than the long-term, `r num_survey_years`-year mean (Figure \ref{fig:ssm_other_ducks}). In `r params$year`, species comprising "other ducks" were: ring-necked duck (`r comp_other_ducks$ring_necked_duck`), hooded merganser (`r comp_other_ducks$hooded_merganser`), common merganser (`r comp_other_ducks$common_merganser`), northern shoveler (`r comp_other_ducks$shoveler`), goldeneye (`r comp_other_ducks$goldeneye`), gadwall (`r comp_other_ducks$gadwall`), and ruddy duck (`r comp_other_ducks$ruddy`). 

### Total Ducks

The `r params$year` population estimate for **all breeding ducks** is **`r all_results$mean` (95% CI: [`r all_results$lower`; `r all_results$upper`])**. This estimate is `r all_results$deviation_previous_year` `r all_results$direction_previous_year` compared to the previous year and is `r all_results$deviation_long_term_mean` `r all_results$direction_long_term_mean` than the long-term, `r num_survey_years`-year mean (Figure \ref{fig:ssm_total_ducks}). As wetland and migratory wildlife species, ducks readily adjust behavior to weather and water conditions. As a result, variations in population estimates and breeding behavior from year to year are expected. While these annual variations draw considerable interest, particularly from duck hunters, looking at the long-range trends is most important for conservation purposes. Wisconsin is fortunate to have a substantial breeding population of ducks supported by mallards, wood ducks, blue-winged teal and others, which each taking advantage of a unique composition of wetland and nesting habitats. Overall, wood duck numbers appear to be increasing, blue-winged teal numbers in Wisconsin are highly variable and should be interpreted in the context of high continental populations, while mallard numbers have shown stabilizing to slight declines as described above. Current Wisconsin landscape conditions combined with a good rain/snowfall year appear capable of supporting total breeding duck numbers of 450,000--550,000. It appears that the Wisconsin landscape will provide good duck production in `r params$year`. Continentally, habitat conditions have been good and duck breeding populations high for several years. However, both habitat and duck numbers are beginning to show some declines as the US and Canadian prairies appear to be in a drying phase. The thousands of lakes and abundant permanent water in Wisconsin help to moderate the impact of drier wetland years by providing stable duck habitat here. The most significant change in current habitat conditions is the ongoing loss of grassland nesting cover in Wisconsin and across the U.S. prairies as a result of changes in federal farm policy. Continued efforts to protect wetland and grassland habitat will be needed to sustain current duck population levels. 

### Canada Geese

Based on the most recent harvest derivations, the proportion of the Wisconsin Canada goose harvest that consists of temperate breeding (formerly 'giant') Canada geese is about 60%, with most of those birds representing Canada geese that breed in Wisconsin [@dooley2017canada]. This proportion indicates the continued importance of in-state breeding Canada geese in our overall fall harvest. The `r params$year` population estimate for **Canada geese is `r cago_results$mean` (95% CI: [`r cago_results$lower`; `r cago_results$upper`])**. This estimate is `r cago_results$deviation_previous_year` `r cago_results$direction_previous_year` than the previous year's modeled estimate and is `r cago_results$deviation_long_term_mean` `r cago_results$direction_long_term_mean` than the long-term, `r num_survey_years`-year mean (Figure \ref{fig:ssm_cago}).

The long-term trend in goose numbers suggests a continued, gradual increase in their population. Field staff indicated that Canada goose nesting and brood production was excellent across the state and we expect numbers similar to recent years going into the fall hunting season. 

While the spring breeding waterfowl survey is designed primarily for ducks, it serves as a fairly reliable index of change in the Canada goose population. Temperate breeding Canada geese represent a positive resource for some Wisconsin residents, but they also represent a problem for other residents in cities and on farms where increasing populations have caused conflicts. Fortunately, there are solutions to the problems that resident geese cause. In Wisconsin, the management strategy for these geese is two-fold: 1) Manage the overall population through hunter harvest, and 2) Address property- or community-specific problems with professionally-guided, integrated management. The early goose hunting season, with a harvest of over 90% local birds during a two-week period, continues to be an important part of our management strategy. The early season now comprises roughly one third of the Canada geese harvested statewide each fall. In addition, site-specific Canada goose control measures are implemented in urban and agricultural areas to mitigate nuisance goose problems. We continue to adapt harvest strategies, banding plans, nuisance goose programs and survey strategies as the breeding population of giant Canada geese increases and expands across Wisconsin (Table \ref{tab:current_breeding_estimate_table}; Figure \ref{fig:ssm_cago}). A continued interest in the early September Canada goose hunting season is important to the overall management of this population. 

\newpage

# Literature Cited

<div id="refs"></div>

\newpage

# Tables

```{r tables, child = here::here("scripts/rmarkdown/final_report_tables.Rmd")}
```

\newpage

# Figures

```{r figures, child = here::here("scripts/rmarkdown/final_report_figures.Rmd")}
```

